### 2010-03-31 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: proofread Universe of inductive types

\section{A universe of simple data-types}
\label{sec:universe-desc}

\begin{wstructure}
<- Why starting with simple data-types
    <- For pedagogical purposes
        <- Data-types as we know them every day
        /> Target dependent types
    -> Cut down version of Induction Recursion
        -> Presentation evolves independently as we go to dependent types
\end{wstructure}

In this section, we describe an implementation of inductive types, as
we know them in simply-typed languages such as Haskell or OCaml. This
choice is motivated by pedagogical considerations: by describing the
layman data-types in our setting, we build on our intuition of
data-types we are familiar with. In particular, we do not aim at
describing a model for generics in simply-typed functional
language. Our goal is and remains to discuss data-types as we find
them in dependently-typed languages.

Our proposal is inspired by, but less involved than, Dybjer and
Setzer's closed formulation of
induction-recursion~\cite{dybjer:general-ir, dybjer:axiom-ir,
  dybjer:ir-initial-algebra, dybjer:iir}. In the context of simple
data-types, the full power of induction-recursion is not
necessary. Therefore, we adopt a stripped-down version of it. As we
extend the scope of our universe of data-types, our presentation will
diverge from induction-recursion. We shall compare both approaches in
Section~\ref{sec:discussion}.

@@

In dependently-typed language, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature, depending on the sensibility of the author.
The notation $\blue{\Sigma}_{\V{x} : \V{A}} (\V{B}\: \V{x})$ stresses the interpretation that
$\Sigma$-types are a generalisation of sums over arbitrary arities.  Where
simply-typed languages have finite sums, dependently-typed languages have
sums indexed by any set.

On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}} (\V{B}\:\V{x})$
emphasises that $\Sigma$-types can be read as a generalisation of products,
where the second component can depend on the first one. When simply-typed
languages pack data into non-dependent tuples, dependently-types languages
tolerate this non-dependent usage. However, they also give the rather novel
ability for data to influence types of further data.

\begin{wstructure}
<- Data-types in the simply-typed world
    -> "sums-of-product"
        <- Sum of constructors
        <- Product of arguments
<- Data-types in the dependently-typed world
    -> "sigmas-of-sigmas"
    /> Need ability to manipulate these sigmas
        -> Define a Code for data-types
        -> Together with a sigma-based Interpretation
\end{wstructure}

In the simply-typed world, the essence of data-type definition can be
summed up by the expression \emph{sum-of-product}. A data-type is defined
by a finite sum -- a choice -- of constructors, each of them is
composed of a product -- a tuple -- of arguments. This form governs
the grammar of data-type definition in languages such as Haskell or
OCaml.

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throws away information when we
crucially need it. In the realm of data-types, things become even
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a sea of $\Sigma$s, our data-types have little
chance of survival. 
\note{Ridiculous tentative of poetry. I was really tired when I wrote that.}

@@

For the structure of our data-types to be known, we crucially need the
ability to name them. \note{Pursuing this crappy analogy. Solving the type
presentation/representation debate would help solving that point.} To this
end, we will use a standard technique in dependently-typed programming: we
construct an ad-hoc universe. This technique dates back to Martin-L\"of
definition of type theory~\cite{martin-lof:itt}. Since then, it has been
fruitful as a programming 
technique~\cite{benke:universe-generic-prog, oury:power-of-pi}
\note{Any idea of other papers?}. 
We refer the reader to the Agda tutorial~\cite{norell:agda-tutorial} for a
pedagogical presentation of universe construction. 
\note{Is that really a motivating motivation for the usage of universe 
construction?}
\note{pwm: Surely ($\EnumU$, $\EnumT$) is a useful example of universes too.
Even if it doesn't show off their full power?}

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we somehow define the syntax of a
language. However, as such, a system of codes is useless as it lacks a
semantics. Instead of equipping the universe of codes with some
computational behaviour, we pragmatically chose to \emph{interpret}
these codes back into the standard type theory. Hence, codes act only
as labels, while the type theory provides the computational
machinery. \note{Is that a good description?}

Codes being simple labels, we have the ability to inspect them, hence
taking advantage of structural information. Being able to inspect
them, we are therefore free to compute over them: deriving new codes
from previous ones, or even new functions on them. In several
occasions, we will have the opportunity to witness the power of
universes.

\begin{wstructure}
<- Justification of the code 
    <- [both figures]: cannot be read separately
    <- Mimic the standard grammar of data-types description
        -> Just as we already know it
        <- '\Sigma for making sigmas-of-sigmas
        <- 'indx for exhibiting the functoriality
            -> For recursive arguments
        <- '1 for end of description
\end{wstructure}

We propose to embed inductive types as a universe in our dependent
type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of data-types definitions in
simply-typed languages. Hence, we have a $\DSigma{\!}{\!}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the data-types is
introduced by $\DIndx{\!}$. When we tie the knot with a fix-point, the
hole left open by $\DIndx{\!}$ will be turned into a standard
recursive argument. We notice that this functoriality appears in the
type of $\descop{\V{D}}{\!}$ itself, for a given data-type definition
$\V{D}$. This corresponds to the object part of the functor described
by $\V{D}$.

@@

To give some intuition on this universe of descriptions, we now turn
to some examples. For obvious pedagogical reasons, we will manually
build these descriptions. However, it should be clear that, in
practice, these definitions can be automatically constructed from a
Haskell-like $\data$ definition.

@@

Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
give a choice between the $\NatZero$ and $\NatSuc{\!}$ constructors. In
the $\NatZero$ case, we reach the end of the description: there is no
useful payload in that case. In the $\NatSuc{\!}$ case, we left one hole
open for the recursive argument, and we close the description.

@@

\begin{wstructure}
<- Tagged description
    <- Form TDesc = List (UId x Desc) [equation]
    <- Follow usual sums-of-product presentation of data-type
        <- Finite set of constructors
        <- Then whatever you want
    -> Any Desc data-type can be turned into this form
        -> No loss of expressive power
        /> Guarantee a ``constructor form''
\end{wstructure}

From the examples above, we observe that data-types are defined by a
$\DSigma{\!}{\!}$ which first argument is a finite set of
constructors. The descriptions fitting into this pattern are called
\emph{tagged} description. Formally, we have:

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its codes, $\DSigma{\!}{\!}$,
$\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
theory. Hence, in an implementation of this type theory, we need to
extended it with these constructors and their typing rules. Although
it makes no conceptual difference, the code of description should
rather be read as a specification. We promise the existence of such
objects, satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

Whereas this definition is amply sufficient in the simply-typed world,
it comes short of its promises in the dependently-typed realm. Indeed,
this type is not dependent, being restricted to a notion of
\emph{iteration}. As this operator will be the corner stone of any
inductive definition, we are better off sharpening it before use.

A full account of such generic induction schema can be found in Benke
et al.~\cite{benke:universe-generic-prog,
  dybjer:ir-initial-algebra}. In the following, we convey the
intuition behind this schema by a step-by-step construction of the
dependent eliminator from the catamorphism. The first step consists in
turning $\V{T}$ into a predicate $\Bhab{\V{P}}{\Mu{\V{D}} \To \Set}$,
hence introducing a dependency on the descriptions. Consequently, we
strengthen the algebra $\descop{\V{D}}{\V{T}} \To \V{T}$ by a similar
process. We start with turning the arrow into a dependent arrow, hence
offering $\Bhab{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}$ to further
analysis.

Then, we have to translate the notion of algebra in our dependent
setting. This is achieved by the predicate $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$
(Fig.~\ref{fig:all-predicates}) that, intuitively, states that
$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Based on that
knowledge, we ought to be able to prove that $\V{P}$ holds for
$\Con{\V{xs}}$. We close our elimination principle by taking an argument
$x$ in $\Mu{\V{D}}$ and claiming that the motive $\V{P}$ is realised on $\V{x}$.

@@

@@

At this stage, we are still left with implementing this
$\F{induction}$ principle. For space reason, we refer the reader to
our technical report for the actual code. Having this dependent
catamorphism, we are empowered with the ability to $\emph{compute}$
over descriptions. Note that this induction principle is the first
manifestation of a generic operation over descriptions. Thanks to
$\F{induction}$, any data-type we define automatically comes equipped
with an induction principle.

@@

\begin{wstructure}
<- Extending type propagation
    <- Data-type declaration turns into definitions
        -> Straightforward translation to Desc
        -> Creation of a variable referring to the structure
    <- Labelled Mu
        /> Just mention the possibility of labelling, no details required
        -> For the user, objects have names rather than Mu of codes
    <- Push Mu to an applied name [figure]
        -> Direct integration into the type propagation machinery
    -> Coded presentation is practical
        <- The user never see a code
\end{wstructure}

So far, we have introduced a universe of description. We have shown
how to manipulate these codes to build data-types. We have also
presented a fix-point construction and its associated dependent
catamorphism. However, one could argue that programming with these
objects is not practical.

Whereas simply-typed languages feature a convenient grammar for
defining data-types and manipulating them, our proposal could be seen
as a step backward. In our setting, the developer would have to write
a description instead of a sugared data-type definition. Then, to
recognise an element of $\Nat$ for instance, she would have to
identify $\Mu{}$ of a description as indeed the fix-point of the
pattern functor of natural numbers. Finally, she would have to write
codes instead of the natural data-type constructor.

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite sets, the
cure lies in rationalising this information. Let us first tackle the
issue of data-type declaration. The grammar of data-types we are used
to can be straightforwardly translated to the sigmas-of-sigmas
paradigm. Hence, a data-type definition can be turned into the
corresponding description. Then, at the level of definitions, we
associate the name of the data-type to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to address
the second issue. Namely, the developer would rather be presented a
name than a fix-point of a pattern functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-defined name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

Regaining the ability to write data-type constructors is simply a
matter of type propagation. We therefore extend the type-checking
framework of Figure~\ref{fig:type-checking} with the following
inference rule:

@@

Hence, in the high-level expression language, the developer writes a
familiar type constructor, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
look-up the tag and turn the constructor form into the appropriate
code.

This concludes our presentation of the universe of descriptions. In
this section, we have seen how data-types from the simply-typed world
can be expressed in our dependently-typed setting. By relying on a
universe construction technique, we benefit from the ability to
inspect and manipulate these codes. This opens some appealing
opportunities for generic programming. Finally, we have demonstrated
that programming with codes is practical. To this end, we have
extended the type propagation machinery. In this setting, we can
define, interact, and build data-types as transparently as in the
simply-typed world.

>>>>

\section{A universe of inductive datatypes}
\label{sec:universe-desc}

\begin{wstructure}
<- Why starting with simple datatypes
    <- For pedagogical purposes
        <- datatypes as we know them every day
        /> Target dependent types
    -> Cut down version of Induction Recursion
        -> Presentation evolves independently as we go to dependent types
\end{wstructure}

In this section, we describe an implementation of inductive types, as
we know them in ML-like languages. This choice is motivated by
pedagogical considerations: by describing the layman datatypes in our
setting, we build on our intuition of datatypes we are familiar
with. In particular, we do not aim at describing a model for generics
in simply-typed functional language. Our goal is and remains to
discuss datatypes as we find them in dependently-typed languages.

Our proposal is inspired by, but less involved than, Dybjer and
Setzer's closed formulation of
induction-recursion~\cite{dybjer:general-ir, dybjer:axiom-ir,
  dybjer:ir-initial-algebra, dybjer:iir}. For inductive datatypes, the
full power of induction-recursion is not necessary. Therefore, we
adopt a stripped-down version of it. As we broaden the scope of our
universe of datatypes, our presentation will diverge from
induction-recursion. We shall compare both approaches in
Section~\ref{sec:discussion}.

@@

In dependently-typed languages, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature, depending on the sensibility of the
author.  The notation $\blue{\Sigma}_{\V{x} : \V{A}} (\V{B}\: \V{x})$
stresses the interpretation that $\Sigma$-types are a generalisation
of sums over arbitrary arities.  Where simply-typed languages have
finite sums, dependently-typed languages have sums indexed by any set.

On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}} (\V{B}\:\V{x})$
emphasises that $\Sigma$-types can be read as a generalisation of products,
where the second component can depend on the first one. When simply-typed
languages pack data into non-dependent tuples, dependently-typed languages
tolerate this non-dependent usage. However, they also give the rather novel
ability for data to influence types of further data.

\begin{wstructure}
<- datatypes in the simply-typed world
    -> "sums-of-product"
        <- Sum of constructors
        <- Product of arguments
<- datatypes in the dependently-typed world
    -> "sigmas-of-sigmas"
    /> Need ability to manipulate these sigmas
        -> Define a Code for datatypes
        -> Together with a sigma-based Interpretation
\end{wstructure}

In the simply-typed world, datatype definitions are often referred to
as \emph{sum-of-product}. A datatype is defined by a finite sum -- a
choice -- of constructors, each of them is composed of a product -- a
tuple -- of arguments. This form governs the grammar of datatype
definition in languages such as Haskell or ML.

To model these datatypes, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. Beyond this
intuition, it should be clear that a direct encoding of datatypes
through raw $\Sigma$-types is not a viable option. Indeed, just as for
finite sets, encoding throws away information when we crucially need
it. Moreover, to program with these datatypes, one would need an
induction principle. An encoding will be unable to deliver such
operation~\cite{geuvers:induction-not-derivable}.

@@

While sigmas-of-sigmas form a convenient \emph{semantics} for
datatypes, we have seen that it makes a poor \emph{syntax}. To
decouple these two notions, we use a standard technique in
dependently-typed programming: we construct an ad-hoc universe. This
technique dates back to Martin-L\"of definition of type
theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
programming technique~\cite{benke:universe-generic-prog,
  oury:power-of-pi} \note{Any idea of other papers?}.  We refer the
reader to the Agda tutorial~\cite{norell:agda-tutorial} for a
pedagogical presentation of universe construction.  \note{Is that
  really a motivating motivation for the usage of universe
  construction?}  \note{pwm: Surely ($\EnumU$, $\EnumT$) is a useful
  example of universes too.  Even if it doesn't show off their full
  power?}

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we define the syntax of a language. However,
as such, a system of codes is useless as it lacks a semantics. Instead
of equipping the universe of codes with some computational behaviour,
we pragmatically chose to \emph{interpret} these codes back into the
standard type theory. Hence, codes act only as labels, while the type
theory provides the computational machinery. \note{Is that a good
  description?}

Codes being simple labels, we have the ability to inspect them, hence
regaining structural information. Being able to inspect them, we are
therefore free to compute over them: deriving new codes from previous
ones, or even new functions on them. In several occasions, we will
have the opportunity to witness the power of universes.

\begin{wstructure}
<- Justification of the code 
    <- [both figures]: cannot be read separately
    <- Mimic the standard grammar of datatypes description
        -> Just as we already know it
        <- '\Sigma for making sigmas-of-sigmas
        <- 'indx for exhibiting the functoriality
            -> For recursive arguments
        <- '1 for end of description
\end{wstructure}

We propose to embed inductive types as a universe in our dependent
type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of datatypes definitions in
simply-typed languages. Hence, we have a $\DSigma{\!}{\!}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the datatypes is
introduced by $\DIndx{\!}$. When we tie the knot with a fix-point, the
hole left open by $\DIndx{\!}$ will be turned into a standard
recursive argument. We notice that this functoriality appears in the
type of $\descop{\V{D}}{\!}$ itself, for a given datatype definition
$\V{D}$. This corresponds to the object part of the functor described
by $\V{D}$.

@@

To give some intuition on this universe of descriptions, we now turn
to some examples. For obvious pedagogical reasons, we will manually
build these descriptions. However, it should be clear that, in
practice, these definitions can be automatically constructed from an
ML-like $\data$ definition.

@@

Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
give a choice between the $\NatZero$ and $\NatSuc{\!}$
constructors. In the $\NatZero$ case, we reach the end of the
description: there is no useful payload in that case. In the
$\NatSuc{\!}$ case, we leave one hole open for the recursive argument,
and we close the description.

@@

\begin{wstructure}
<- Tagged description
    <- Form TDesc = List (UId x Desc) [equation]
    <- Follow usual sums-of-product presentation of datatype
        <- Finite set of constructors
        <- Then whatever you want
    -> Any Desc datatype can be turned into this form
        -> No loss of expressive power
        /> Guarantee a ``constructor form''
\end{wstructure}

From the examples above, we observe that datatypes are defined by a
$\DSigma{\!}{\!}$ which first argument is a finite set of
constructors. The descriptions fitting into this pattern are called
\emph{tagged} description. Formally, we have:

@@

For convenience, we have taken for granted the existence of $\Desc$,
as presented in Figure~\ref{fig:desc_universe}. In particular, we have
considered its codes, $\DSigma{\!}{\!}$, $\DIndx{\!}$, and $\DUnit$ as
type formers, extending the basic type theory. Hence, in an
implementation of this type theory, we need to extended it with these
constructors and their typing rules. Although it makes no conceptual
difference, the code of description should rather be read as a
specification. We promise the existence of such objects, satisfying
the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

Whereas this definition is amply sufficient in the simply-typed world,
it comes short of its promises in the dependently-typed realm. Indeed,
this type is not dependent, being restricted to a notion of
\emph{iteration}. As this operator will be the corner stone of any
inductive definition, we are better off sharpening it before use.
Following \citet{benke:universe-generic-prog}, we adopt the following
definition:

@@

Where, intuitively, $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$ states that
$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its formal
definition is presented in Figure~\ref{fig:all-predicates}.  We refer
the reader to our technical report for the implementation of
$\F{induction}$. Having this dependent catamorphism, we are empowered
with the ability to $\emph{compute}$ over descriptions. Note that this
induction principle is the first manifestation of a generic operation
over descriptions. Thanks to $\F{induction}$, any datatype we define
automatically comes equipped with an induction principle.

@@

@@

\begin{wstructure}
<- Extending type propagation
    <- datatype declaration turns into definitions
        -> Straightforward translation to Desc
        -> Creation of a variable referring to the structure
    <- Labelled Mu
        /> Just mention the possibility of labelling, no details required
        -> For the user, objects have names rather than Mu of codes
    <- Push Mu to an applied name [figure]
        -> Direct integration into the type propagation machinery
    -> Coded presentation is practical
        <- The user never see a code
\end{wstructure}

We have now enough machinery to build and manipulate inductive
types. In a word, we ought be able to \emph{program} over these
datatypes.  However, one could argue that programming with these
objects is not practical.

Whereas simply-typed languages feature a convenient grammar for
defining datatypes and manipulating them, our proposal could be seen
as a step backward. In our setting, the developer would have to write
a description instead of a sugared datatype definition. Then, to
recognise an element of $\Nat$ for instance, she would have to
identify $\Mu{}$ of some coded description as indeed the fix-point of
the pattern functor of natural numbers. Finally, she would have to
write codes instead of datatype constructors.

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite sets, the
cure lies in rationalising this information. Let us first tackle the
issue of datatype declaration. The grammar of datatypes we are used
to can be straightforwardly translated to the sigmas-of-sigmas
paradigm. Hence, a datatype definition can be turned into the
corresponding description. Then, at the level of definitions, we
associate the name of the datatype to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.

Doing so, we seal datatype definition behind a convenient
abstraction. We take advantage of this abstraction barrier to address
the second issue. Namely, the developer would rather be presented a
name than a fix-point of a pattern functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-defined name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

Regaining the ability to write datatype constructors is simply a
matter of type propagation. We therefore extend the type-checking
framework of Figure~\ref{fig:type-checking} with the following
inference rule:

@@

Hence, in the high-level expression language, the developer writes a
familiar type constructor, applied to a telescope of arguments. During
type propagation, this expression will be pushed to a $\Mu{}$ type of
a tagged description. It is then straightforward to look-up the tag
and turn the constructor form into the appropriate code.

This concludes our presentation of the universe of descriptions. In
this section, we have seen how datatypes from the simply-typed world
can be expressed in our dependently-typed setting. By relying on a
universe construction technique, we benefit from the ability to
inspect and manipulate these codes. This opens some appealing
opportunities for generic programming. Finally, we have demonstrated
that programming with codes is practical. To this end, we have
extended the type propagation machinery. In this setting, we can
define, interact, and build datatypes as transparently as in the
simply-typed world.
